---
title: " Statistical Linear Models Assigment "
author: "Patrick Dhatemwa, Ian Mulindwa, Cynthia Musangi, Francis Gitau"
date: "`r Sys.Date()`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objective : Fit a linear model with mpg as the dependent variable and wt and hp as the independent variables.

Loading the mtcars dataset using the code below:
``` {r attach mtcars}
                    #  data(mtcars)
                    # attach(mtcars)
```

## For a better understanding of the data, we view the first 5 rows.

```{r mtcars}

head(mtcars)

```

## To fit the model, we start by defining varibales

```{r Fitting the model.}

#----------------------------------------
#          QUESTION 1 (a)
#----------------------------------------


# Set mpg as the independent variable. (Y)
Y <- (mtcars$mpg)
Y


```

## We now go ahead to define the dependent variable. (X1 [wt], and X2 [hp])

After defining the dependent variables, we then go ahead to find the transpose of the matrix of dependent variables to meet our objective.

```{r independent variables, eval=TRUE}


# Set wt as the dependent variable. (X)
data(mtcars)
attach(mtcars)
X <- cbind(wt, hp)
X_t <- t(X) # find the transpose of X.
X_t # Call the function.


```

##Step 1:  Fitting the linear model.
## OLS formulation.

```{r OLS formulation}

N <- X_t%*%X
N # Call the function

```


##Step 2: Obtain the inverse of the matrix using the solve function.
```{r}
# Obtain the inverse of the matrix resulting from the product of the X matrix and its transpose.
solve(N)
```

## Step 3: Now we obtain the matrix obtained from the product of X-transpose and the Y-matrix
``` {r}
f <- X_t%*%Y
f
```

 
# Note: according to OLS, B = [([X_t][X])^-1] * [(X_t)*(Y)]
# But according to our code, that will be Solve(N) * f

## Step 4 : Final OLS formulation.
# -------------------------------
```{r}
B = solve(N)%*%f
B # Call B
```
# -------------------------------

## Step 5 : The linear model.


```{r}
model <- lm(mpg ~ wt + hp, data = mtcars)
show(model)
summary(model)
```

---

i)	P- value :  9.109e-12
This P-value is less than 0.05, and therefore we can say that our model is significant.

ii)	Interpretation of coefficients.
Bo - The minimum value of mpg when hp and (wt) are zero. (i.e. 37.23) which is the minimum level of fuel efficiency.

B1 – holding horsepower constant, a unit increase in (wt) decreases fuel efficiency (mpg) by 3.87783 units. Meaning if the weight of the car increases by about 1000 pounds, full efficiency drops by about 3.88 mpg

B2 - holding weight constant, a unit increase in horse power decreases (mpg) by 0.03177 units 

iii)	a) R-squared Value (0.8268) : This means that 82.68% of the variation in the model can be explained by (wt) and (hp).

  b) Adjusted R-squared value (0.8148) : This means that 81.48% of the variation in the model can still be explained by (wt) and (hp) meaning that the model is still very significant since the difference between the R-squared and the adjusted R-squared is not very big.
  
---


## B) REGRESSION ASSUMPTIONS.
```{r}
# i)	 Check linearity and interpret the findings.

plot(model$fitted.values, resid(model),
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted Values")
abline(h = 0, col = "red")
```


### Assumption 1: LINEARITY

We are checking to see if residuals are randomly scattered around zero and not having an obvious pattern or curve which violates linearity.
According to the above plot, linearity is not completely violated.


### Assumption 2 : Check independence of errors.

```{r, eval=TRUE}
#install.packages("lmtest")
#install.packages("car")

library(lmtest)
library(car)

dwtest(model)
```

Here we are checking whether residuals are independent of one another.


Hypotheses
H0 : Errors are independent

H1 : Errors are auto-correlated.

Interpretation

If p-value > 0.05, we fail to reject H0. Independence is acceptable.
If p-value < 0.05, there may be autocorrelation.

But since p-value < 0.05, we conculde that there may be auto-correlation.



### Assumption 3 : Check constant variance (homoscedasticity).
```{r}
plot(model$fitted.values, resid(model),
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Homoscedasticity Check")
abline(h = 0, col = "red")
```
Here we want the residuals to have roughly equal spread across fitted values.

# Interpretation
If residuals spread evenly, homoscedasticity is satisfied.
If the residuals form a funnel shape, variance is not constant.

And according to the above plot, homoscedasticity is satisfied, since the residuals spread evenly.

### Assumption 4 : Check the normality of residuals.
```{r normality_of_residuals}
hist(resid(model),
     main = "Histogram of Residuals",
     xlab = "Residuals")

qqnorm(resid(model))
qqline(resid(model), col = "red")

shapiro.test(resid(model))
```
### Assumption 5 : Check for multicollinearity.
```{r multicollinearity}
library(car)
vif(model)
```
Here we are checking whether weight, (wt) and horse power, (hp) are too highly correlated.

#Interpretation

Common rule:
VIF < 5      acceptable
VIF 5 to 10  concerning
VIF > 10     serious multicollinearity

If VIF values are below 5, multicollinearity is not a serious problem.

Since both values for wt, and hp are < 5, the test outcome is accessible.

### C) Perform Case Diagnostics.
## (i)Standard residuals
```{r}
std_res <- rstandard(model)
std_res
```

```{r}
# To identify possible outliers
which(abs(std_res) > 2)
```

Common rule:
|standardized residual| > 2  possible outlier
|standardized residual| > 3  serious outlier
These identify observations whose actual mpg is unusually far from predicted mpg.

Car	Standardized  -------   Residual
Chrysler Imperial -------	  2.3545
Fiat 128	        -------   2.3319
Toyota Corolla	  -------   2.3786

## (ii) Cook's Distance
```{r}
cook <- cooks.distance(model)
cook

plot(cook, type = "h",
     main = "Cook's Distance",
     ylab = "Cook's Distance")
abline(h = 4 / nrow(mtcars), col = "red")

# To identify influential cases
which(cook > 4 / nrow(mtcars))

```

## Interpretation.

Cook’s distance checks whether one observation strongly affects the regression model.
Rule:
Cook′s Distance > 4/n
 
Since n=32:  4/32 = 0.125
Any case above 0.125 may be influential.

According to our output,
The following are influetial to out liear regression model, meaning removing would significantly affect our model.

Car               ------  Cook’s Distance
Chrysler Imperial ------	0.4236
Fiat 128	        ------  0.1574
Toyota Corolla	  ------  0.2087
Maserati Bora	    ------  0.2720

Which exceed the threshold of 0.125

Cook’s Distance:
0.4236
This is the largest influence in the model.
Meaning, the Chrysler Imperial substantially affects the regression estimates. Removing it would noticeably change the fitted regression model.

## (iii) Leverage
```{r}
lev <- hatvalues(model)
lev

plot(lev, type = "h",
     main = "Leverage Values",
     ylab = "Leverage")
abline(h = 2 * 3 / nrow(mtcars), col = "red")

# To identify high leverage cases.
which(lev > 2 * 3 / nrow(mtcars))
```

Interpretation
High leverage observations have unusual predictor values. They may not be outliers in mpg, but their wt and hp combinations are unusual.

## (iv) Covarince Ratio
```{r}
cov_ratio <- covratio(model)
cov_ratio

# We now check for influental cases
n <- nrow(mtcars)
p <- 3

lower <- 1 - (3 * p / n)
upper <- 1 + (3 * p / n)

lower
upper

which(cov_ratio < lower | cov_ratio > upper)
```

Covariance ratio checks whether removing a case substantially changes the precision of the model estimates.
 
So the acceptable range is approximately:
0.71875 to 1.28125
Cases outside this range may be influential. namely: 
1) Cadillac Fleetwood
2) Lincoln Continental
3) Fiat 128
4) Toyota Corolla
5) Ford Pantera L
6) Maserati Bora
































